---
title: "Part 1: Detecting Olive Oil Fraud Using Unsupervised Learning"
author: "Derick Malavi"
format: 
  html:
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: style.css
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"  # Adjust margins
    output: rmarkdown::github_document
---

### Introduction

Fraudulent adulteration of **Extra-Virgin Olive Oil (EVOO)** with cheaper oils continues to be a significant concern due to its economic and health implications. Traditional methods for detecting adulteration, such as Gas Chromatography-Mass Spectrometry (GC-MS), are effective but suffer from drawbacks—they are destructive, time-consuming, and costly. As a result, novel non-destructive methods are being explored, including **Near-Infrared Hyperspectral Imaging (NIR-HSI)**.

NIR-HSI is a technique that combines imaging with spectroscopy, capturing both spatial and spectral information across a wide range of wavelengths. This makes it an ideal candidate for detecting subtle changes in EVOO that indicate adulteration. The richness of the spectral data presents a challenge: the data sets are typically large and complex, requiring advanced techniques to reduce dimensionality and reveal underlying patterns. Unsupervised learning methods, such as **Principal Component Analysis (PCA)** and **K-Means clustering**, are crucial tools for addressing this challenge ( [Malavi et al., 2023](https://www.mdpi.com/2304-8158/12/3/429)).

### Principal Component Analysis (PCA)

PCA is a powerful tool for dimensionality reduction, especially in the context of hyperspectral data, where hundreds of spectral bands are captured. By transforming the original variables into a smaller set of uncorrelated principal components, PCA allows us to identify the most significant spectral features contributing to variance in the dataset. This technique is particularly useful in exploring the relationships between different oils and detecting outliers or adulteration patterns without prior labeling of the data.

In this study, PCA was applied to the spectral data obtained from NIR-HSI of pure and adulterated EVOO samples. The goal was to reduce the dimensionality of the data while preserving the most critical information. This not only facilitates visualization but also prepares the data for further unsupervised classification through K-Means clustering.

### K-Means Clustering

K-Means clustering is an unsupervised learning technique used to group data points into clusters based on similarity. When applied to the transformed data from PCA, K-Means can group EVOO samples based on their spectral similarities, allowing for the identification of clusters that correspond to pure and adulterated oils. The advantage of K-Means lies in its simplicity and efficiency, making it suitable for large datasets, such as those generated by NIR-HSI.

In this project, K-Means clustering was performed on the principal components derived from the PCA of NIR-HSI data. The clustering process helped to separate the pure EVOO samples from the adulterated ones, providing a preliminary classification without the need for labeled training data. This unsupervised approach serves as a foundational analysis before moving to supervised classification models.

### Objectives of Part 1

-   **Dimensionality Reduction**: Utilize PCA to reduce the complexity of the hyperspectral data while retaining the most important spectral features for differentiating pure and adulterated EVOO.

-   **Unsupervised Classification**: Perform K-Means clustering on the PCA-transformed data to classify EVOO samples based on their spectral characteristics, identifying potential adulteration.

-   **Exploratory Data Analysis**: Use PCA and K-Means to uncover hidden patterns in the data, providing insights into how adulteration affects the spectral properties of EVOO.

## Dataset

The dataset contains pure samples of extra-virgin olive oil and those adulterated/mixed with different cheaper oils, including safflower, corn, sesame, sunflower, canola, and soybean oils, in concentration ranges of 0-20% (m/m). The samples were analyzed by GC-MS, FTIR, Raman, UV-Vis, and HSI to determine if these methods could identify them as genuine or not.

**Load Installed Packages from the R Library for the Analysis**

```{r}
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret)    
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(janitor)  
  library(FactoMineR)
  library(MASS)
  library(factoextra)
  library(rmarkdown)
  library(knitr) 
  library(officedown)
  library(quarto)
  library(gtsummary)
  library(pander)
  library(tinytex)
  library(kernlab)}))
```

**Load and Inspect Hyperspectral Imaging Data**

```{r}
#Load HSI spectra data
hsi<-read_excel("HSI.xlsx")
#Check dimensions
dim(hsi)
#We have 183 observations and 228 variables
```

```{r}
#Check a few of the column names
colnames(hsi[,c(1:4)])
#Check for any missing values
anyNA(hsi)# There are no missing values
```

```{r}
#Considering that we are conducting a binary classification, we will remove some columns
hsi<-hsi[,-c(1,3)]
table(hsi$class_1)
#convert class_1 to a factor
hsi$class_1<-as.factor(hsi$class_1)
#There are two classes: The oils are either pure/authentic or adulterated
```

```{r}
#Check whether the data is normalized. We want a value between 0 and 1
print(paste('The max value is', max(hsi[,-c(1:4)]), 
            'and the min value is', min(hsi[,-c(1:4)])))
```

**Load and Inspect Raman Spectroscopy Data**

```{r}
#Load Raman spectra data
raman<-read_excel("Raman.xlsx")
#Check dimensions
dim(raman)
#We have 183 observations and 1404 variables
```

```{r}
#Bind the data to have the same columns as HSI data
raman<-cbind(hsi[,c(1,2)],raman[,-c(1:3)])
colnames(raman[,c(1:3)])#check whether the changes have been effected
table(raman$class_1)#Check the class distribution
class(raman$class_1)#ensure class_1 is a factor
```

```{r}
# Define the normalization function to have values of 0 and 1
min_max_normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}
raman<-min_max_normalize(raman[,-c(1:2)])
print(paste('The max value is', max(raman[,-c(1:2)]), 'and the min value is', min(raman[,-c(1:2)])))
anyNA(raman)
dim(raman)
```

**Load and Inspect FTIR Spectroscopy Data**

```{r}
#Load FTIR spectra data
ftir<-read_excel("FTIR.xlsx")
#Check dimensions
dim(ftir)
#We have 183 observations and 919 variables
```

```{r}
#Bind the data to have the same columns as HSI data
ftir<-cbind(hsi[,c(1,2)],ftir[,-c(1:3)])
colnames(ftir[,c(1:3)])#check whether the changes have been effected
table(ftir$class_1)#Check the class distribution
class(ftir$class_1)#ensure class_1 is a factor
print(paste('The max value is', max(ftir[,-c(1:2)]), 
            'and the min value is', min(ftir[,-c(1:2)])))#The data is OK
dim(ftir)
```

**Load and Inspect UV-Vis Spectroscopy Data**

```{r}
#Load Uv-Vis spectra data
uv_vis<-read_excel("UVVIS.xlsx")
#Check dimensions
dim(uv_vis)
```

```{r}
#Bind the data to have the same columns as HSI data
uv_vis<-cbind(hsi[,c(1,2)],uv_vis[,-c(1:4)])
colnames(uv_vis[,c(1:3)])#check whether the changes have been effected
table(uv_vis$class_1)#Check the class distribution
class(uv_vis$class_1)#ensure class_1 is a factor
print(paste('The max value is', max(uv_vis[,-c(1:2)]), 
            'and the min value is', min(uv_vis[,-c(1:2)])))#The data is OK
dim(uv_vis)
```

```{r}
# Define the normalization function to have values of 0 and 1
min_max_normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}
uv_vis<-min_max_normalize(uv_vis[,-c(1:2)])
print(paste('The max value is', max(uv_vis[,-c(1:2)]), 
            'and the min value is', min(uv_vis[,-c(1:2)])))
anyNA(uv_vis)
dim(uv_vis)
#There are 183 observations and 121 covariates
```

**Load and Inspect GC-MS Data**

```{r}
#Load GC-MS data
gc_ms<-read_excel("GC_MS.xlsx")
#Check dimensions
dim(gc_ms)
gc_ms$class_1<-factor(gc_ms$class_1)#convert to factor
```

```{r}
# Define the normalization function to have values of 0 and 1
min_max_normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}
gc<-min_max_normalize(gc_ms[,-c(1:2)])
print(paste('The max value is', max(gc), 'and the min value is', min(gc)))
anyNA(gc_ms)
gc_ms<-cbind(gc_ms[,c(1,2)],gc)
```

## Unsupervised Learning

### Exploratory Data Analysis by Principal Component Analysis (PCA)

Principal Component Analysis (PCA) is vital for managing the complexity of high-dimensional spectral data from HSI, FTIR, and Raman spectroscopy. It reduces dimensionality, filters noise, extracts key features, improves computational efficiency, and addresses multicollinearity, resulting in more effective and insightful scientific analysis.

#### Run PCA

```{r}
hsi_pca<-PCA(hsi[,-c(1,2)],graph = F)#HSI data
hsi_pca$eig[1:10,]#extracting the first 5 components' eigenvalues
```

```{r}
raman_pca<-PCA(raman[,-c(1,2)],graph = F)#Raman data
raman_pca$eig[1:10,]#extract the first 5 components' eigenvalues
```

```{r}
ftir_pca<-PCA(ftir[,-c(1,2)],graph = F)#FTIR data
ftir_pca$eig[1:10,]#extract the first 5 components' eigenvalues
```

```{r}
uv_vis_pca<-PCA(uv_vis[,-c(1,2)],graph = F)#UV-Vis data
uv_vis_pca$eig[1:10,]#extract the first 5 components' eigenvalues
```

```{r}
gc_ms_pca<-PCA(gc_ms[,-c(1,2)],graph = F)#GC-MS data
gc_ms_pca$eig[1:7,]
```

#### Scree Plots

**HSI**

```{r}
s1<-fviz_eig(hsi_pca, addlabels = TRUE, ylim = c(0, 90),xlim=c(1,5), main = 'HSI Scree Plot',barfill = "chocolate1",hjust = 0.5,ncp = 9,ggtheme = theme_bw(),xlab = "PCs", ylab = "% variance")+
  theme(plot.title = element_text(hjust = 0.5))+
   theme(panel.grid = element_blank())
print(s1)
```

**Raman**

```{r}
s2<-fviz_eig(raman_pca, addlabels = TRUE, ylim = c(0, 60),xlim=c(1,5),main = 'Raman Scree Plot',barfill = "greenyellow",hjust = 0.5,ncp = 20,ggtheme = theme_bw(),xlab = "PCs",ylab = "% variance")+
  theme(plot.title = element_text(hjust = 0.5))+
   theme(panel.grid = element_blank())
print(s2)
```

**FTIR**

```{r}
s3<-fviz_eig(ftir_pca, addlabels = TRUE, ylim = c(0, 90),xlim=c(1,5),main = 'FTIR Scree Plot',barfill = "dodgerblue1",hjust = 0.5,ncp = 7,ggtheme = theme_bw(),xlab = "PCs", ylab = "% variance")+
  theme(plot.title = element_text(hjust = 0.5))+
   theme(panel.grid = element_blank())
print(s3)
```

**Uv-Vis**

```{r}
s4<-fviz_eig(uv_vis_pca, addlabels = TRUE, ylim = c(0, 90),xlim=c(1,5),main = 'UV-Vis Scree Plot',barfill = "goldenrod4",hjust = 0.5,ncp = 15,ggtheme = theme_bw(),xlab = "PCs", ylab = "% variance")+
  theme(plot.title = element_text(hjust = 0.5))+
  theme(panel.grid = element_blank())
print(s4)
```

```{r}
#Patch together the scree plots
gridExtra::grid.arrange(s1,s2,s3,s4, nrow =2)
```

**GC-MS**

```{r}
s5<-fviz_eig(gc_ms_pca, addlabels = TRUE, ylim = c(0, 40),main = 'GC-MS',barfill = "lightblue",hjust = 0.5,ncp = 8,ggtheme = theme_bw(),xlab = "PCs", ylab = "% variance")+
  theme(plot.title = element_text(hjust = 0.5))+
   theme(panel.grid = element_blank())

print(s5)
```

-   As observed from the data, **more than 80% of the variation** in the data from each of the spectroscopic technique can be explained by the the first 3 PCs.

### Visualization of PC Scores

Different principal components will be examined to visualize any patterns from our data. The next step will be to create a new data frame for each technique to be used in subsequent analysis.

```{r}
#HSI Data
hsi_new<-as.data.frame(hsi_pca$ind$coord) #Extract the PCs
colnames(hsi_new)<-c("PC_1","PC_2", "PC_3","PC_4","PC_5")
hsi_new<-cbind(hsi[,c(1,2)],hsi_new)#Bind with the dependent variables
head(hsi_new)
```

```{r}
#Raman Data
raman_new<-as.data.frame(raman_pca$ind$coord) #Extract the PCs
colnames(raman_new)<-c("PC_1","PC_2", "PC_3","PC_4","PC_5")
raman_new<-cbind(hsi[,c(1,2)],raman_new)#Bind with the dependent variables
head(raman_new)
```

```{r}
#FTIR Data
ftir_new<-as.data.frame(ftir_pca$ind$coord) #Extract the PCs
colnames(ftir_new)<-c("PC_1","PC_2", "PC_3","PC_4","PC_5")
ftir_new<-cbind(hsi[,c(1,2)],ftir_new)#Bind with the dependent variables
head(ftir_new)
```

```{r}
#Uv_Vis Data
uvvis_new<-as.data.frame(uv_vis_pca$ind$coord) #Extract the PCs
colnames(uvvis_new)<-c("PC_1","PC_2", "PC_3","PC_4","PC_5")
uvvis_new<-cbind(hsi[,c(1,2)],uvvis_new)#Bind with the dependent variables
head(uvvis_new)
```

```{r}
#GC-MS Data
gc_new<-as.data.frame(gc_ms_pca$ind$coord) #Extract the PCs
colnames(gc_new)<-c("PC_1","PC_2", "PC_3","PC_4","PC_5")
gc_new<-cbind(gc_ms[,c(1,2)],gc_new)#Bind with the dependent variables
head(gc_new)
```

#### **PC Plots**

```{r}
# HSI PC Plot
p1 <- hsi_new %>% 
  ggplot(mapping = aes(x = PC_1, y = PC_2, shape = class_1, color = perc_adulter)) +
  geom_point() + 
  labs(x = "PC1 (79.8%)", y = "PC2 (10.9%)",title = "HSI PC Plot", shape = "Oil type", color = "Percent Adulteration") +
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(color = 'black', size = 10),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
     legend.position = "none") +
  scale_color_gradient(low = "#000000", high = "red") +
  stat_ellipse(aes(group = class_1), 
               level = 0.95, 
               geom = "polygon", alpha = 0.2,
               color = 'black', linewidth = 0.6)
```

```{r}
#Raman Plot
p2 <- raman_new %>% 
  ggplot(mapping = aes(x = PC_2, y = PC_3, shape = class_1, color = perc_adulter)) +
  geom_point() + 
  labs(x = "PC2 (17.1%)", y = "PC3 (10.7%)",title = "Raman PC Plot", shape = "Oil type", color = "Percent Adulteration") +
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    axis.text.x = element_text(color = 'black', size = 10),
    panel.grid = element_blank(),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6)) +
  scale_color_gradient(low = "#000000", high = "red") +
  stat_ellipse(aes(group = class_1), 
               level = 0.95, 
               geom = "polygon", alpha = 0.2,
               color = 'black', linewidth = 0.6)
```

```{r}
#FTIR Plot
p3 <- ftir_new %>% 
  ggplot(mapping = aes(x = PC_2, y = PC_3, shape = class_1, color = perc_adulter)) +
  geom_point() + 
  labs(x = "PC2 (10.9%)", y = "PC3 (7.0%)",title = "FTIR PC Plot", shape = "Oil type", color = "Percent Adulteration") +
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    axis.text.x = element_text(color = 'black', size = 10),
    panel.grid = element_blank(),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
     legend.position = "none") +
  scale_color_gradient(low = "#000000", high = "red") +
  stat_ellipse(aes(group = class_1), 
               level = 0.95, 
               geom = "polygon", alpha = 0.2,
               color = 'black', linewidth = 0.6)
```

```{r}
#Uv_Vis Plot
p4 <- uvvis_new%>% 
  ggplot(mapping = aes(x = PC_1, y = PC_2, shape = class_1, color = perc_adulter)) +
  geom_point() + 
  labs(x = "PC1 (74.5%)", y = "PC2 (11.7%)",title = "Uv_Vis PC Plot", shape = "Oil type", color = "Percent Adulteration") +
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(color = 'black', size = 10),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6)) +
  scale_color_gradient(low = "#000000", high = "red") +
  stat_ellipse(aes(group = class_1), 
               level = 0.95, 
               geom = "polygon", alpha = 0.2,
               color = 'black', linewidth = 0.6)
```

#### **Patch the PC Plots together**

```{r}
suppressWarnings(suppressMessages(library(gridExtra)))
grid.arrange(p1,p2,p3,p4, nrow = 2)
```

```{r}
#GC-MS Plot
p5 <- gc_new %>% 
  ggplot(mapping = aes(x = PC_1, y = PC_2, shape = class_1, color = perc_adulter)) +
  geom_point() + 
  labs(x = "PC1 (39.3%)", y = "PC2 (18.5%)",title = "GC-MS PC Plot", shape = "Oil type", color = "Percent Adulteration") +
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(color = 'black', size = 10),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6)) +
  scale_color_gradient(low = "#000000", high = "red") +
  stat_ellipse(aes(group = class_1), 
               level = 0.95, 
               geom = "polygon", alpha = 0.2,
               color = 'black', linewidth = 0.6)
#Display plot
p5
```

### PCA Insights

-   PCA results indicate interesting patterns. Although the separation does not appear to be very clear, authentic olive oil tends to separate from adulterated olive oils, especially with HSI, UV-Vis, and GC-MS. PCA, however, demonstrates weakness in discerning oils adulterated at different levels, hence the need for additional supervised algorithms.

### PC variable contributions

This section investigates the contribution of different variables to the variation in principal components (PCs). By analyzing the loadings of each variable on the principal components, we can determine which variables have the most significant impact on the observed patterns in the data. This analysis helps to identify key features that drive the separation of samples in the PCA plot.

```{r}
#HSI
h1<-fviz_contrib(hsi_pca, choice = "var", top = 5,axes = 1, sort.val = 'desc', fill = "olivedrab")
h2<-fviz_contrib(hsi_pca, choice = "var", top = 5,axes = 2, sort.val = 'desc', fill = "cadetblue")
grid.arrange(h1,h2, nrow = 1)
```

```{r}
#Raman
r1<-fviz_contrib(raman_pca, choice = "var", top = 5,axes = 1, sort.val = 'desc', fill = "#E7B800")
r2<-fviz_contrib(raman_pca, choice = "var", top = 5,axes = 2, sort.val = 'desc', fill = "#00AFBB")
grid.arrange(r1,r2, nrow = 1)
```

```{r}
#FTIR
f1<-fviz_contrib(ftir_pca, choice = "var", top = 5,axes = 1, sort.val = 'desc', fill = "tan4")
f2<-fviz_contrib(ftir_pca, choice = "var", top = 5,axes = 2, sort.val = 'desc', fill = "cornflowerblue")
grid.arrange(f1,f2, nrow = 1)
```

```{r}
#Uv-Vis
uv1<-fviz_contrib(uv_vis_pca, choice = "var", top = 5,axes = 1, sort.val = 'desc', fill = "thistle")
uv2<-fviz_contrib(uv_vis_pca, choice = "var", top = 5,axes = 2, sort.val = 'desc', fill = "powderblue")
grid.arrange(uv1,uv2, nrow = 1)
```

```{r}
#GC-MS
gc1<-fviz_contrib(gc_ms_pca, choice = "var", top = 5,axes = 1, sort.val = 'desc', fill = "chocolate")
gc2<-fviz_contrib(gc_ms_pca, choice = "var", top = 5,axes = 2, sort.val = 'desc', fill = "gray58")
grid.arrange(gc1,gc2, nrow = 1)
```

### K-Means Clustering

K-means clustering is an unsupervised machine learning algorithm used to partition a dataset into **K distinct, non-overlapping groups** (or clusters) based on feature similarity. It aims to minimize the variance within each cluster and maximize the variance between clusters.

-   Let us find the number of [**clusters**]{.underline} based on silhoutte method

```{r}
# optimal number of clusters for HSI
hsi_clust<-fviz_nbclust(hsi[,-c(1:2)], kmeans, method = "silhouette", k.max=10)
print(hsi_clust)
```

```{r}
# optimal number of clusters for Raman
raman_clust<-fviz_nbclust(raman, kmeans, method = "silhouette", k.max=10)
print(raman_clust)
```

```{r}
#optimal number of clusters for FTIR
ftir_clust<-fviz_nbclust(ftir[,-c(1,2)], kmeans, method = "silhouette", k.max=10)
print(ftir_clust)
```

```{r}
#optimal number of clusters for UV-Vis
uvvis_clust<-fviz_nbclust(uv_vis, kmeans, method = "silhouette", k.max=10)
plot(uvvis_clust)
```

```{r}
#optimal number of clusters for gc-ms
gc_clust<-fviz_nbclust(gc, kmeans, method = "silhouette", k.max=10)
print(gc_clust)
```

-   Then **let us perform k-means clustering with the optimal number of clusters**

```{r}
#HSI k-means analysis and plots

hsi_kmeans <- kmeans(hsi[,-c(1,2)],2)
cluster<-  hsi_kmeans$cluster
hsi_k_data <-cbind(hsi_new,cluster)
hsi_k_data$cluster<-as.factor(hsi_k_data$cluster)

hsi_k_data %>% 
  ggplot(mapping = aes(x = PC_1, y = PC_2, color = cluster, shape = class_1)) +
  geom_point() + 
  labs(x = "PC1", y = "PC2",title = "HSI k-means cluster Plot")+
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(color = 'black', size = 10),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.position = "right")+
  scale_color_manual(values = c("blue", "red"))
```

```{r}
#Raman k-means analysis and plotting 

raman_kmeans <- kmeans(raman,3)
cluster<-  raman_kmeans$cluster
raman_k_data <-cbind(raman_new,cluster)
raman_k_data$cluster<-as.factor(raman_k_data$cluster)

raman_k_data %>% 
  ggplot(mapping = aes(x = PC_1, y = PC_2, color = cluster, shape = class_1)) +
  geom_point() + 
  labs(x = "PC1", y = "PC2",title = "Raman k-means cluster Plot")+
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(color = 'black', size = 10),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.position = "right")+
  scale_color_manual(values = c("blue", "red","black"))
```

```{r}
#FTIR k-means analysis and plotting

ftir_kmeans <- kmeans(ftir[,-c(1,2)],3)
cluster<-  ftir_kmeans$cluster
ftir_k_data <-cbind(ftir_new,cluster)
ftir_k_data$cluster<-as.factor(ftir_k_data$cluster)

ftir_k_data %>% 
  ggplot(mapping = aes(x = PC_1, y = PC_2, color = cluster, shape = class_1)) +
  geom_point() + 
  labs(x = "PC1", y = "PC2",title = "FTIR k-means cluster Plot")+
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(color = 'black', size = 10),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.position = "right")+
  scale_color_manual(values = c("blue", "red","black"))
```

```{r}
#UV-Vis k-means analysis and plotting

uvvis_kmeans <- kmeans(uv_vis,2)
cluster<-  uvvis_kmeans$cluster
uvvis_k_data <-cbind(uvvis_new,cluster)
uvvis_k_data$cluster<-as.factor(uvvis_k_data$cluster)

uvvis_k_data %>% 
  ggplot(mapping = aes(x = PC_1, y = PC_2, color = cluster, shape = class_1)) +
  geom_point() + 
  labs(x = "PC1", y = "PC2",title = "UV-Vis k-means cluster Plot")+
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(color = 'black', size = 10),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.position = "right")+
  scale_color_manual(values = c("blue", "red"))
```

```{r}
#GC-MS k-means analysis and plotting

gc_kmeans <- kmeans(gc,2)
cluster<-  gc_kmeans$cluster
gc_k_data <-cbind(gc_new,cluster)
gc_k_data$cluster<-as.factor(gc_k_data$cluster)

gc_k_data %>% 
  ggplot(mapping = aes(x = PC_1, y = PC_2, color = cluster, shape = class_1)) +
  geom_point() + 
  labs(x = "PC1", y = "PC2",title = "GC-MS k-means cluster Plot")+
  theme_bw() +
  theme(
    panel.border = element_rect(color = 'black', fill = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(color = 'black', size = 10),
    axis.text.y = element_text(color = 'black', size = 10), 
    aspect.ratio = 1,
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.position = "right")+
  scale_color_manual(values = c("blue", "red"))
```

-   Based on the findings from k-means clustering, an overlap of clusters for the two expected groups is observed. Unsupervised learning, therefore, is not sufficient for separating pure EVOO from adulterated samples, highlighting the need for supervised classification.

### Summary

This section of the project demonstrates how unsupervised learning techniques, particularly PCA and K-Means clustering, can be applied to the complex spectral data obtained from NIR-HSI. These methods provide a powerful way to reduce data complexity and uncover patterns that may not be immediately apparent, laying the groundwork for further classification using supervised machine learning models.

### Way Forward to Part 2: Supervised Classification Using Machine Learning

Now that **unsupervised learning** techniques like PCA and K-Means clustering have helped reduce the dimensionality of the Near-Infrared Hyperspectral Imaging (NIR-HSI) data and revealed underlying patterns, the next step is to leverage **supervised classification** methods to develop robust models that can distinguish between **pure and adulterated Extra-Virgin Olive Oil (EVOO)**.
